\Title

### 2020-12-09 manuscript/manuscript.tex
Shorten and summarize basic EQL theory in methods
I was a bit unsure about the explanations for the theory behind the
equivalent layer that we were presenting. It felt a bit too short with
some important parts missing about the upward continuation equation and
Green's identities. But at the same time it was feeling too long for a
paper (considering we already have to explain the gradient boosting
scheme). So I figured it might be good to cheat and start from the
Cordell 1992 and skip the deductions of why we can use 1/r. That's been
done in books and we have the citation for Cordell so we don'

\documentclass[twocolumn]{article}
\usepackage[left=0.7in,right=0.7in,top=1in,bottom=1in]{geometry}
\setlength{\columnsep}{2\columnsep}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[round,authoryear,sort]{natbib}
\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\usepackage{fancyhdr}
\usepackage{multirow}
\usepackage{booktabs}

@@

Many variants of the equivalent source technique have been proposed, often
attempting to obtain faster or more accurate solutions.
The key factors that vary between them are: (i) the type of source, (ii)
the location of the sources, and (iii) the solution strategy.
The type of source is most commonly a point mass for gravity or dipole for
magnetics \citep[e.g.,~][]{vonfrese1981, silva1986, mendonca1994, siqueira2017}.
However, right-rectangular prisms \citep[e.g.,][]{barnes2011, jirigalatu2019,
li2020} and tesseroids \citep{bouman2016} have also been used successfully.
In fact, even point sources with a simple inverse distance function, instead of
actual gravity or magnetic fields, can be used as
equivalent sources \citep{cordell1992}.
The sources are often distributed on a regular grid at a constant depth
\citep[e.g.,~][]{leao1989, barnes2011, oliveira2013}
or placed beneath each data point \citep[e.g.,~][]{cordell1992, siqueira2017}.
The model is usually estimated through damped least-squares, which imposes a
heavy computational load when the number of data points is large (e.g.,
airborne and satellite surveys).
To reduce the computational load, \citet{mendonca1994} built the solution
iteratively by incorporating one data point at a time using the ``equivalent
data concept''.
\citet{leao1989} processed the input data using a moving window, only fitting the
data inside the window and predicting observations at the center of the window.
\citet{li2010} and \citet{barnes2011} apply different operations to generate a
sparse representation of the sensitive matrix (respectively, wavelet
compression and quadtree discretization), which significantly improves the
speed of the least-squares solution.
\citet{oliveira2013} parametrized the equivalent layer as a piecewise bivariate
polynomial function, reducing the number of parameters in the solution.
\citet{siqueira2017} developed an iterative solution in which the sensitivity
matrix is transformed into a diagonal matrix with constant terms through the
``excess mass criterion''.
\citet{jirigalatu2019} applied the Gauss-FFT method to speed up the forward
modelling operations and solved the least-squares problem using steepest
descent to avoid calculating the Hessian matrix and solving linear systems.
Most existing methods solve under-determined problems, requiring a much larger
number of equivalent sources than the number of data points.
Furthermore, many of the optimizations proposed are also complex to implement
in a computer program, limiting their wider adoption.
In the present study,
we propose a two-way method for reducing the computational load needed to apply
the equivalent sources technique: first by reducing the number of equivalent
sources through a \emph{block-averaging} strategy and then simplifying the
fitting process through a \emph{gradient boosting} algorithm.
The first strategy consists in
dividing the survey area into horizontal blocks and assign one source
to each block at the median horizontal location of the data points.
For airborne surveys, which are oversampled along tracks, this can greatly
reduce the size of the inverse problem while retaining the quality of
interpolation.
The second algorithm simplifies the process of fitting a very large number of
equivalent sources by defining sets of fewer sources, whose coefficients are
fitted iteratively while minimizing a given residue.
Through tests on synthetic data, we show that: (i) the \emph{block-averaged}
sources achieve the same accuracy as a traditional equivalent layer while using
a fraction of the number of sources, and (ii) the \emph{gradient boosting}
algorithm greatly reduces the computational memory required to fit very large
datasets without scarifying predictions' accuracy.
Finally these methods are used to interpolate a 1.7 million gravity data
measurements around Australia onto a regular grid.

@@

\section{The equivalent sources technique}

The Green's equivalent layer theorem states that any harmonic potential field
generated by any three dimensional physical-property distribution can by
reproduced by a continuous two dimensional surface distribution.
This introduces an ambiguity on the distribution that generates
a potential field: multiple physical-property distributions can generate the
same potential field.
This represents an obstacle for determining the true shape and
physical-property distribution that generates any measured potential field.
Nevertheless this ambiguity can also be exploited for developing pre-processing
techniques like interpolating potential field data.

Approximating the continuous distribution by a finite and discrete density
distribution is the basis of the \emph{equivalent sources technique}
\citep{dampney1969}.
The most common approach is to make use of a set of point sources (the
\emph{equivalent sources}) located beneath the observation points.

A classical approach for applying the equivalent sources technique would be to
propose a set of point sources that generate the same harmonic field that has
been measured.
Lets consider a set of gravity acceleration measurements $g_z$ on points
$\{\mathbf{p}_1, \ldots, \mathbf{p}_N\}$ and a set of point sources with mass
$\{m_1, \ldots, m_M\}$ located on points
$\{\mathbf{q}_1, \ldots, \mathbf{q}_M\}$.
The point sources must generate the same gravity field as the one measured:

\begin{equation}
    g_z(\mathbf{p}) = \mathbf{A} \mathbf{m},
    \label{eq:linear-problem-g}
\end{equation}

\noindent where $g_z(\mathbf{p})$ is the vector containing the measured
field on the points $\mathbf{p}$, $\mathbf{m}$ is the vector containing the
masses of the point sources and $\mathbf{A}$ a matrix whose
elements can be obtained as the vertical component of the gravitational
acceleration measured on $\mathbf{p}$ due to the point mass on $\mathbf{q}$:

\begin{equation}
    a_{ij} = G \frac{(z_i - z_j)}{|\mathbf{p}_i - \mathbf{q}_j|^3},
\end{equation}

\noindent where
$G = \SI{6.674e-11}{\cubic\meter\per\kilogram\per\second\squared}$ is the
universal gravitational constant, $z_i$ and $z_j$ are the vertical components
of the $\mathbf{p}_i$ and $\mathbf{q}_j$ points, respectively.

By solving equation~\ref{eq:linear-problem-g} we can obtain the values of
$\mathbf{m}$ that better reproduce the measured field.
This source distribution along with the fitted masses can be used to predict
gravity values on any point where the potential field has not been measured.

\subsection{Generalized equivalent sources}

It's possible to generalize the equivalent sources technique to any harmonic
field through a Green's function problem, regardless the nature of the field.

By definition, any harmonic field $V$ must satisfy the Laplace's equation:

\begin{equation}
    \nabla^2 V = 0.
\end{equation}

Considering as boundary condition that $V$ is equal to zero at infinity, then
its fundamental solution $\psi$ (also known as Green's function or radial basis
function) must satisfy \citep[p.~210]{vladimirov1979}:

\begin{equation}
    \nabla_\mathbf{p}^2 \psi(\mathbf{p}, \mathbf{q}) =
        - 4\pi \delta(\mathbf{p}, \mathbf{q}),
\end{equation}

\noindent where $\delta(\mathbf{p}, \mathbf{q})$ is the Dirac's delta function
and $\nabla_\mathbf{p}$ symbolize the Nabla operator applied on the coordinates
$\mathbf{p}$.

The fundamental solution of any harmonic field can be obtained as follows
\citep[p.~37]{blakely1995}:

\begin{equation}
    \psi(\mathbf{p}, \mathbf{q})
    =
    \int
    \frac{\delta(\mathbf{p}, \mathbf{q})}{|\mathbf{p} - \mathbf{q}|}
    \textrm{d}v
    =
    \frac{1}{|\mathbf{p} - \mathbf{q}|}.
\end{equation}

This shows that the Green's function associated with any harmonic field is
equal to the inverse of the Euclidean distance between the evaluation and the
source points.
Thus any harmonic field $V$ can be expressed as \citep[p.~37]{blakely1995}:

\begin{equation}
    V(\mathbf{p})
    =
    \int \psi(\mathbf{p}, \mathbf{q}) \rho(\mathbf{q}) d\mathbf{q}
    =
    \int \frac{\rho(\mathbf{q})}{|\mathbf{p} - \mathbf{q}|}  d\mathbf{q},
\end{equation}

\noindent where $\rho(\mathbf{q})$ is the density distribution of a certain
physical-property evaluated on the point $\mathbf{q}$. Following
\citet{dampney1969}, this density distribution can be approximated by a finite
and discrete set of point sources located on $\{\mathbf{q}_1, \ldots,
\mathbf{q}_M\}$:

\begin{equation}
    \rho(\mathbf{q}) =
        \sum\limits_{j=1}^{M} c_j \delta(\mathbf{q} - \mathbf{q_j}),
\end{equation}

\noindent where $c_j$ is a coefficient related to the point source located on
$\mathbf{q}_j$, with the same units as the physical-property.
Then the harmonic field $V$ can be approximated by:

\begin{equation}
    V(\mathbf{p})
    =
    \sum\limits_{j=1}^{M} \frac{c_j}{|\mathbf{p} - \mathbf{q}_j|}.
    \label{eq:eql-field}
\end{equation}

In case we have measured the value of the harmonic field on points
$\{\mathbf{p}_1, \ldots, \mathbf{p}_N\}$, we can adjust the values of the $c$
coefficients so they satisfy:

\begin{equation}
    V(\mathbf{p}_i)
    =
    \sum\limits_{j=1}^{M} \frac{c_j}{|\mathbf{p}_i - \mathbf{q}_j|}
    \quad \forall i=1,\ldots,N.
\end{equation}

This system of linear equations can be written in an more algebraic way:

\begin{equation}
    \mathbf{d} = \mathbf{A} \mathbf{c}
    \label{eq:linear-problem}
\end{equation}

\noindent where $\mathbf{d}$ is a vertical vector containing the measured
values of $V$ on the observation points, $\mathbf{c}$ a vertical vector
containing the $c$ coefficients and $\mathbf{A}$ is the \emph{Jacobian matrix},
whose elements are defined as the Green's function evaluated on each pair of
points $\mathbf{p}$ and $\mathbf{q}$:

\begin{equation}
    a_{ij}
    =
    \psi(\mathbf{p}_i, \mathbf{q}_j)
    =
    \frac{1}{|\mathbf{p}_i - \mathbf{q}_j|}
\end{equation}

By solving equation~\ref{eq:linear-problem} we can obtain the values of
$\mathbf{c}$ that better reproduce the measured field.
Then they can be used to predict the value of the harmonic field on any other
point where no measurement has been taken by evaluating
equation~\ref{eq:eql-field}.

>>>>

\documentclass[twocolumn]{article}
%\usepackage[left=0.7in,right=0.7in,top=1in,bottom=1in]{geometry}
\setlength{\columnsep}{2\columnsep}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[round,authoryear,sort]{natbib}
\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\usepackage{fancyhdr}
\usepackage{multirow}
\usepackage{booktabs}

@@

Many variants of the equivalent source technique have been proposed, often
attempting to obtain faster or more accurate solutions.
The key factors that vary between them are: (i) the type of source, (ii)
the location of the sources, and (iii) the solution strategy.
The type of source is most commonly a point mass for gravity or dipole for
magnetics \citep[e.g.,~][]{vonfrese1981, silva1986, mendonca1994, siqueira2017}.
However, right-rectangular prisms \citep[e.g.,][]{barnes2011, jirigalatu2019,
li2020} and tesseroids \citep{bouman2016} have also been used successfully.
In fact, even point sources with a simple inverse distance function, instead of
actual gravity or magnetic fields, can be used as
equivalent sources \citep{cordell1992}.
The sources are often distributed on a regular grid at a constant depth
\citep[e.g.,~][]{leao1989, barnes2011, oliveira2013}
or placed beneath each data point \citep[e.g.,~][]{cordell1992, siqueira2017}.
The model is usually estimated through damped least-squares, which imposes a
heavy computational load when the number of data points is large (e.g.,
airborne and satellite surveys).
To reduce the computational load, \citet{mendonca1994} built the solution
iteratively by incorporating one data point at a time using the ``equivalent
data concept''.
\citet{leao1989} processed the input data using a moving window, only fitting the
data inside the window and predicting observations at the center of the window.
\citet{li2010} and \citet{barnes2011} apply different operations to generate a
sparse representation of the sensitive matrix (respectively, wavelet
compression and quadtree discretization), which significantly improves the
speed of the least-squares solution.
\citet{oliveira2013} parametrized the equivalent layer as a piecewise bivariate
polynomial function, reducing the number of parameters in the solution.
\citet{siqueira2017} developed an iterative solution in which the sensitivity
matrix is transformed into a diagonal matrix with constant terms through the
``excess mass criterion''.
\citet{jirigalatu2019} applied the Gauss-FFT method to speed up the forward
modelling operations and solved the least-squares problem using steepest
descent to avoid calculating the Hessian matrix and solving linear systems.
Most existing methods solve under-determined problems, requiring a much larger
number of equivalent sources than the number of data points.
Furthermore, many of the optimizations proposed are also complex to implement
in a computer program, limiting their wider adoption.

In the present study,
we propose a two-way method for reducing the computational load needed to apply
the equivalent sources technique: first by reducing the number of equivalent
sources through a \emph{block-averaging} strategy and then simplifying the
fitting process through a \emph{gradient boosting} algorithm.
The first strategy consists in
dividing the survey area into horizontal blocks and assign one source
to each block at the median horizontal location of the data points.
For airborne surveys, which are oversampled along tracks, this can greatly
reduce the size of the inverse problem while retaining the quality of
interpolation.
The second algorithm simplifies the process of fitting a very large number of
equivalent sources by defining sets of fewer sources, whose coefficients are
fitted iteratively while minimizing a given residue.
Through tests on synthetic data, we show that: (i) the \emph{block-averaged}
sources achieve the same accuracy as a traditional equivalent layer while using
a fraction of the number of sources, and (ii) the \emph{gradient boosting}
algorithm greatly reduces the computational memory required to fit very large
datasets without scarifying predictions' accuracy.
Finally these methods are used to interpolate a 1.7 million gravity data
measurements around Australia onto a regular grid.

@@

\section{Methodology}

\subsection{The equivalent sources technique}

Here, we will follow the ``generalized equivalent sources'' of
\citet{cordell1992} and assume that any harmonic function $d(\mathbf{p})$ can
be approximated by a sum of $M$ discrete point source effects

\begin{equation}
    d(\mathbf{p})
    =
    \sum\limits_{j=1}^{M} \frac{c_j}{\left\lVert \mathbf{p} - \mathbf{q}_j
    \right\rVert} \ ,
    \label{eq:eql-forward}
\end{equation}

\noindent in which
$\mathbf{p}$ and $\mathbf{q}_j$ are position vectors in a 3D Cartesian space,
$c_j$ is a scalar coefficient related to the point source located at
$\mathbf{q}_j$,
and $\lVert \cdot \rVert$ is the vector norm.

In case we have measured the value of the harmonic field on $N$ points
$\{\mathbf{p}_1\ \mathbf{p}_2\ \ldots\ \mathbf{p}_N\}$,
we can write a set of $N$ equations of the form

\begin{equation}
    d_i
    =
    \sum\limits_{j=1}^{M} \frac{c_j}{\left\lVert \mathbf{p}_i - \mathbf{q}_j
    \right\rVert}
    \quad \forall i=1,2,\ldots,N
    \ ,
\end{equation}

\noindent which can also be expressed as the following matrix equation

\begin{equation}
    \mathbf{d} = \mathbf{A} \mathbf{c} \ ,
    \label{eq:linear-problem}
\end{equation}

\noindent where $\mathbf{d}$ is a column vector containing the $N$ predicted
values of the field $d_i$ on the observation points $\mathbf{p}_i$,
$\mathbf{c}$ is a column vector containing the $M$ coefficients $c_j$,
and $\mathbf{A}$ is the $N \times M$ Jacobian matrix,
whose elements are

\begin{equation}
    a_{ij} = \frac{1}{\left\lVert\mathbf{p}_i - \mathbf{q}_j\right\rVert}
\end{equation}

For a given set of $N$ observed field values $\mathbf{d}^o$,
we can find a least-squares solution to
equation~\ref{eq:linear-problem} and obtain the values of
$\mathbf{c}$ that best fit the observations.
These values can, in turn, be used to predict the value of the harmonic field
at any other point outside of the sources by evaluating
equation~\ref{eq:eql-forward}.
Griding and upward continuation can thus be achieved by predicting values on
points that fall on a regular grid or at different heights.
